## Getting Started

To begin, copy all the text on this screen. Then log in to https://rstudio.middlebury.edu. In the "File" menu select "New File" then "R Markdown". Add a title like "SOAN 245 Workshop Week 3", include your name in the "Author" box, and click Ok. A default markdown file will open. Delete all the text in the default file below line 7 and paste in the text you copied to your clipboard. Make sure that "R Markdown" is selected in the file type drop down menu in the bottom right corner of this pane. Save the file (in the R menu, not your browser menu!) in your "Home" folder on the R Studio server. 

***Once the R Markdown file is saved, you can access it through the server in the future. You do not have to copy and paste it from the web link each time. The first place to look for it is in the "File" menu (in the R menu, not your browser menu!) under "Recent Files".***

## Once Your File Is Saved

## Loading Data and Packages

The data we are using come from the Billion Prices Project. In the transparent spirit of open science, they make all their data available for download at: http://www.thebillionpricesproject.com/datasets/. Today we are using data from the "Online-Offline Price Comparison" paper which analyzes price data from the following retailers: Best Buy, CVS, Walmart, Target, Lowe's, Macy's, Office Max, or Staples. 

Let's load the packages and the data. (Note: there may be some additional packages you have to install.)

```{r, message = FALSE, warning = FALSE, error = FALSE}
install.packages(c("bitops", "RCurl"))
library(RCurl) #We need this to read the data
library(dplyr) #We need this to clean up the data
library(ggplot2) #We need this for graphics
workshop3_raw_data <- getURL("https://raw.githubusercontent.com/lawrence-midd/soan245/master/bpp.csv")
bpp <- read.csv(text = workshop3_raw_data)
```

You know the dataset loaded correctly if it appears in R Studio's top right pane. To see a list of the variables in our dataset, click the white arrow in the blue circle. To see the data as a table, click the spreadsheet icon. 

## Data Cleaning

Big data can be messy, especially when scraped from the web. We need to clean up a couple things in the dataset before using it. (Note: I cleaned up some of the really messy stuff before making the data available to you, so this looks a little different from data you could download directly from the BPP website.)

Here's the first clean up issue. There are prices in the dataset that are for products that are on sale and prices that are for products that are not on sale. Prices that are on sale are coded with a 1 for the variable `sale_online`, and prices that are not on sale are coded with a 0 for that variable. 

Why would we want to drop the prices that are on sale? Why would we want to keep the prices that are on sale?

To drop all the prices that are on sale, we filter our dataset to only include observations for prices that are *not* on sale. The syntax in the chunk below says:

"Remake the dataset called `bpp` by filtering the existing dataset called `bpp` to only keep the observations where the variable `sale_online` has the value `0`."

```{r}
bpp <- filter(bpp, sale_online==0)
```

How many observations do we have now?

We also want to make sure we are only looking at recent prices. Let's only keep the prices that have been scraped within the previous 7 days. The variable for how many days it has been since the price was scraped is called `datediff`. Using the `filter` function again as we did in the chunk above, how might we only include observations for prices scraped within the previous 7 days?

### REPLACE THIS LINE WITH YOUR CODE ###






How similar do you think the prices will be at the retailers' brick and mortar stores (variable = `price_offline`) and on their own websites (`price_online`)? What are some tools we could use to compare them?

### REPLACE THIS LINE WITH YOUR CODE ###







## Finding Price Differences

Now it's time to calculate the differences betweeen the price for a product if purchased from Amazon and the price for the same product if purchased from the retailer's own website. The function to create a new variable is called `mutate`. We are going to create a new variable called `price_difference_amon` that will take the values of the differences of the prices. We'll use the log of the prices so we can calculate the percentage differences.

```{r}
bpp <- mutate(bpp, price_difference_amon = log(price_amazon) - log(price_online))
```

If this worked correctly, we should now have 21 variable in the `bpp` dataset.

Techical detour: maybe some differences are so small that we should consider the prices from Amazon and the online store to be the same. If the absolute value of the difference is less than 1%, we should recode it as 0. Let's find the absolute value of the difference first:

```{r}
bpp <- mutate(bpp, abs_price_difference_amon = abs(price_difference_amon))
```

And then recode values as appropriate. If the absolute value of the difference is less than .01, we will recode the difference as 0. If the absolute value of the difference is something else, its value remains the value of `abs_price_difference_amon`.

```{r}
bpp <- mutate(bpp, abs_price_difference_amon = ifelse(abs_price_difference_amon<0.01,0,abs_price_difference_amon))
```

Let's take a first look at the distribution of these price differences:

```{r}
boxplot(bpp$abs_price_difference_amon)
```

Some of these seem way too large to be true, which is a sign that there might be some errors in the data. We'll only keep the cases where the difference between the Amazon price and the online price is .5 or les. Do you remember how to `filter` the dataset to only keep some cases?

```{r}
bpp <- filter(bpp, abs_price_difference_amon<=.5)
```

Now let's get percentages to make some of our analyses easier to interpret:

```{r}
bpp <- mutate(bpp, price_difference_amon_pc=price_difference_amon*100)
```

Take a look at a boxplot of the price difference percentages:

### REPLACE THIS LINE WITH YOUR CODE ###





## Analysis

Let's compare the price differences! We'll start by looking at a summary:

```{r}
summary(bpp$price_difference_amon_pc)
```

What is the difference between the negative and positive values?

The summary is helpful, but a figure showing the distribution might be even more helpful. We are going to make a histogram. The first step is to lay out the background for our figure. We create a new object called `diff_plot` (or whatever you want to call it) which will use the `ggplot` package to take the `bpp` dataset and populate the x-axis with the values of the `price_difference_amon_pc` variable.

```{r, message = FALSE}
diff_plot <- ggplot(bpp, aes(x = price_difference_amon_pc))
```

Recall the object name to see what we have so far:

```{r, message = FALSE}
diff_plot
```

Next let's add a histogram to this background layer:

```{r, message = FALSE}
diff_plot + geom_histogram()
```

The histogram above shows the *number* of cases in the y-axis and the percent difference in the x-axis.

It is probably preferable to switch the y-axis to the *percent* of cases rather than the count. We can redo all of the commands in one chunk instead of building up layer by layer:

```{r, message = FALSE}
diff_plot_pc <- ggplot(bpp, aes(x = price_difference_amon_pc, y=..count../sum(..count..)))

diff_plot_pc + geom_histogram()
```

Finally let's add a title and some better axis labels to make our figure cleaner:
  
```{r, message = FALSE}

diff_plot_pc + geom_histogram() + 
  labs(x = "Price difference (percent)", 
       y = "Percent", 
       title = "Price Differences With Amazon.com (U.S. Only)")
```

How does this histogram compare to Figure 2 (page 298) of Cavallo's 2017 paper?


## Price Differences by Product Type

The dataset groups products into four types: Electronics, Home and Appliances, Office Products, Pharmacy and Health, or Mix (meaning a product falls into one or more of the other categories). How would you expect the distribution of price differences to vary by product type?

We can use the `facet_grid()` function to have a separate plot for each value of the `category` variable:

```{r, message = FALSE}

diff_plot_pc + geom_histogram() + 
  labs(x = "Price difference (percent)", 
       y = "Percent", 
       title = "Price Differences With Amazon.com (U.S. Only)") +
  facet_grid(.~ category)

```


## Table of Differences

The histograms are great for giving us a general sense of the distribution. To get some more specific details about how prices compare, we can make a table.

Start by cleaning up some of the data. In the code chunk below, we'll recode all price differences that have an absolute value of 1% or less as being identical.

```{r}
bpp <- mutate(bpp, price_difference_amon_pc = ifelse(abs(price_difference_amon_pc)<=1,0,price_difference_amon_pc))
```

Now let's group the prices into three groups based on whether the prices are identical, Amazon is higher, or Amazon is lower:

```{r}
bpp <- bpp %>% 
  mutate(difference = ifelse(price_difference_amon_pc==0,"Identical",
      ifelse(price_difference_amon_pc>0, "Higher Amazon", "Lower Amazon"))) %>%
  mutate(difference = factor(difference,
      levels = c("Identical", "Higher Amazon", "Lower Amazon")))
```

And finally let's make a proportion table showing the differences:
  
```{r}
round(prop.table(table(bpp$difference)),3)
```

These proportions very closely reproduce the first row of Table 8 (page 299) in the Cavallo (2017) paper.
